---
title: "COVID-19 Risk Factor Data Cleaning"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: html_document
---

## Introduction
This notebook completes the data cleaning work for this project.

## Dependencies
The following packages are needed for this part of the project:

```{r load-dependencies}
# tidyverse
library(dplyr)       # data wrangling
library(readr)       # read and write csv files

# spatial packages
library(sf)          # spatial tools
library(tidycensus)  # download demographic data
library(tigris)      # download geometric data

# other packages
library(here)        # file path management
```

## Geometric Data
### Download
First, we'll download the raw zip code data for the U.S.:

```{r get-zips}
zip <- zctas(year = 2018, class = "sf") %>%
  st_transform(crs = 26915)
```

We'll also download county data for Missouri:

```{r get-counties}
counties <- counties(state = 29, year = 2018, class = "sf") %>%
  filter(COUNTYFP %in% c("183", "189", "510")) %>%
  st_transform(crs = 26915)
```

### Geoprocess
With our data downloaded we can intersect them to identify zips within St. Louis County, St. Louis City, and St. Charles County:

```{r process-zips}
# intersect
intersect <- st_intersection(zip, counties)

# remove stragglers
zip_stl <- filter(zip, GEOID10 %in% intersect$GEOID10) %>%
  filter(GEOID10 >= 63000) %>% # exclude zips in IL
  filter(GEOID10 %in% c(63073, 63055, 63090) == FALSE) %>% # exclude sw zips
  filter(GEOID10 %in% c(63390, 63379, 63362, 63369) == FALSE) %>% # exclude nw zips
  select(GEOID10)
```

With them clean, we'll write them:

```{r write-zips}
zip_stl %>%
  st_transform(crs = 4326) %>%
  st_write(., here("data", "spatial", "zips.geojson"), delete_dsn = TRUE)
```

## Demographic Data
### Downloald Poverty Data
First, we'll calculate poverty rates for each zip code. We need to download the data and rename variables:

```{r}
poverty <- get_acs(table = "B17001", geography = "zcta", output = "wide") %>%
  filter(GEOID %in% zip_stl$GEOID10) %>%
  select(GEOID, B17001_001E, B17001_001M, B17001_002E, B17001_002M) %>%
  rename(
    total = B17001_001E, 
    total_moe = B17001_001M, 
    poverty = B17001_002E, 
    poverty_moe = B17001_002M
  )
```

### Download Age Data
We'll repeat the process for age, but will need to calculate across a lot of columns to figure out the proportion of folks over 65 years in age:

```{r}
age <- get_acs(table = "B01001", geography = "zcta", output = "wide") %>%
  filter(GEOID %in% zip_stl$GEOID10) %>%
  select(GEOID, B01001_001E, B01001_001M, 
         B01001_020E, B01001_021E, B01001_022E, B01001_023E, B01001_024E, B01001_025E,
         B01001_044E, B01001_045E, B01001_046E, B01001_047E, B01001_048E, B01001_049E) %>%
  mutate(over65 = B01001_020E + B01001_021E + B01001_022E + B01001_023E + B01001_024E + B01001_025E +
           B01001_044E + B01001_045E + B01001_046E + B01001_047E + B01001_048E + B01001_049E) %>%
  select(GEOID, B01001_001E, B01001_001M, over65) %>%
  rename(
    total = B01001_001E,
    total_moe = B01001_001M
  )
```

## Mortality Data
We're going to process mortality data both for all adults and only those over 65.

### Cancer
We'll start by processing cancer data:

```{r process-cancer}
# cancer, all
cancer <- read_csv(here("data", "mortality", "raw", "cancer_all.csv"), skip = 2) %>%
  rename(
    GEOID = `Year:`,
    total_cancer = `Total for selection`
  ) %>%
  filter(GEOID %in% c("Statistics:", "Zip / ZCTA") == FALSE) %>%
  slice(1:(n()-6)) %>%
  select(GEOID, total_cancer) %>%
  mutate(total_cancer = as.numeric(total_cancer))

# cancer, over65
cancer65 <- read_csv(here("data", "mortality", "raw", "cancer_over65.csv"), skip = 2) %>%
  rename(
    GEOID = `Year:`,
    total_cancer65 = `Total for selection`
  ) %>%
  filter(GEOID %in% c("Statistics:", "Zip / ZCTA") == FALSE) %>%
  slice(1:(n()-6)) %>%
  select(GEOID, total_cancer65) %>%
  mutate(total_cancer65 = as.numeric(total_cancer65))
```

### COPD
Next, we'll process data on COPD, a respritory aliment:

```{r process-copd}
# copd, all
copd <- read_csv(here("data", "mortality", "raw", "copd_all.csv"), skip = 2) %>%
  rename(
    GEOID = `Year:`,
    total_copd = `Total for selection`
  ) %>%
  filter(GEOID %in% c("Statistics:", "Zip / ZCTA") == FALSE) %>%
  slice(1:(n()-6)) %>%
  select(GEOID, total_copd) %>%
  mutate(total_copd = as.numeric(total_copd))

# copd, over65
copd65 <- read_csv(here("data", "mortality", "raw", "copd_over65.csv"), skip = 2) %>%
  rename(
    GEOID = `Year:`,
    total_copd65 = `Total for selection`
  ) %>%
  filter(GEOID %in% c("Statistics:", "Zip / ZCTA") == FALSE) %>%
  slice(1:(n()-6)) %>%
  select(GEOID, total_copd65) %>%
  mutate(total_copd65 = as.numeric(total_copd65))
```

### Diabetes
Next, we'll process data on mortality due to diabetes:

```{r process-diabetes}
# diabetes, all
diabetes <- read_csv(here("data", "mortality", "raw", "diabetes_all.csv"), skip = 2) %>%
  rename(
    GEOID = `Year:`,
    total_diabetes = `Total for selection`
  ) %>%
  filter(GEOID %in% c("Statistics:", "Zip / ZCTA") == FALSE) %>%
  slice(1:(n()-6)) %>%
  select(GEOID, total_diabetes) %>%
  mutate(total_diabetes = as.numeric(total_diabetes))

# diabetes, over65
diabetes65 <- read_csv(here("data", "mortality", "raw", "diabetes_over65.csv"), skip = 2) %>%
  rename(
    GEOID = `Year:`,
    total_diabetes65 = `Total for selection`
  ) %>%
  filter(GEOID %in% c("Statistics:", "Zip / ZCTA") == FALSE) %>%
  slice(1:(n()-6)) %>%
  select(GEOID, total_diabetes65) %>%
  mutate(total_diabetes65 = as.numeric(total_diabetes65))
```

### Heart Disease
Finally, we'll process data on heart disease mortality:

```{r process-heart-disease}
# heart disease, all
heart <- read_csv(here("data", "mortality", "raw", "heart_disease_all.csv"), skip = 2) %>%
  rename(
    GEOID = `Year:`,
    total_heart = `Total for selection`
  ) %>%
  filter(GEOID %in% c("Statistics:", "Zip / ZCTA") == FALSE) %>%
  slice(1:(n()-6)) %>%
  select(GEOID, total_heart) %>%
  mutate(total_heart = as.numeric(total_heart))

# heart disease, over65
heart65 <- read_csv(here("data", "mortality", "raw", "heart_disease_over65.csv"), skip = 2) %>%
  rename(
    GEOID = `Year:`,
    total_heart65 = `Total for selection`
  ) %>%
  filter(GEOID %in% c("Statistics:", "Zip / ZCTA") == FALSE) %>%
  slice(1:(n()-6)) %>%
  select(GEOID, total_heart65) %>%
  mutate(total_heart65 = as.numeric(total_heart65))
```

### Combine Mortality Data
With all our mortality data complete, we'll combine it together. There are a couple zip codes we don't want to include on our map, so we'll subset those out as well:

```{r combine-mortality}
# join all
left_join(cancer, copd, by = "GEOID") %>%
  left_join(., diabetes, by = "GEOID") %>%
  left_join(., heart, by = "GEOID") %>%
  filter(GEOID %in% c("63010", "63045", "63155") == FALSE) -> mortality_all

# join over 65
left_join(cancer65, copd65, by = "GEOID") %>%
  left_join(., diabetes65, by = "GEOID") %>%
  left_join(., heart65, by = "GEOID") %>%
  filter(GEOID %in% c("63010", "63045", "63155") == FALSE) -> mortality_over65

# clean-up
rm(cancer, cancer65, copd, copd65, diabetes, diabetes65, heart, heart65)
```

## Combine Demographic and Mortality Data
With our mortality data cleaned, we'll combine it with our demographic data:

```{r combine-with-demos}
# join demographics
mortality_all <- left_join(poverty, mortality_all, by = "GEOID")
mortality_over65 <- left_join(age, mortality_over65, by = "GEOID")

# clean-up
rm(age, poverty)
```

Next, we'll calculate totals as well as rates per 1000 residents:

```{r calculate-rates}
# calculate totals and rates
mortality_all %>%
  rowwise() %>%
  mutate(mortality = sum(total_cancer, total_copd, total_diabetes, total_heart, na.rm=TRUE)) %>%
  ungroup() %>%
  mutate(mortality_rate = mortality/total*1000) %>%
  mutate(poverty_rate = poverty/total*1000) %>%
  select(GEOID, total, total_moe, poverty, poverty_moe, poverty_rate, everything()) -> mortality_all

mortality_over65 %>%
  rowwise() %>%
  mutate(mortality = sum(total_cancer65, total_copd65, total_diabetes65, total_heart65, na.rm=TRUE)) %>%
  ungroup() %>%
  mutate(over65_rate = over65/total*1000) %>%
  mutate(mortality_rate = mortality/over65*1000) %>%
  select(GEOID, total, total_moe, over65, over65_rate, everything()) -> mortality_over65
```

Finally, we'll write our cleaned data:

```{r write-clean}
# write data
write_csv(mortality_all, here("data", "mortality", "mortality_all.csv"))
write_csv(mortality_over65, here("data", "mortality", "mortality_over65.csv"))
```

