---
title: "COVID-19 Risk Factor Mapping"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(leaflet)
library(biscale)
library(readr)
library(here)
library(sf)

zips <- st_read(here("data", "spatial", "zips.geojson"), crs = 4326, stringsAsFactors = FALSE) %>%
  rename(GEOID = GEOID10) %>%
  mutate(GEOID = as.numeric(GEOID)) %>%
  filter(GEOID %in% c(63010, 63045, 63155) == FALSE)

all <- read_csv(here("data", "mortality", "mortality_all.csv")) %>%
  left_join(zips, ., by = "GEOID")

over65 <- read_csv(here("data", "mortality", "mortality_over65.csv")) %>%
  left_join(zips, ., by = "GEOID")

rm(zips)
```

## Background
This interactive report contains demographic data drawn from the 2014-2018 American Community Survey's 5-year estimates for Zip Code Tabulation Areas (ZCTA). These are not *exactly* the same as zip codes, but are used to provide standard demographic estimates that roughly correspond to zip code coverage areas. Both the count of individuals living below the federal poverty line and the county of individuals 65 and over were converted to rates per 1,000.

Mortality data are drawn from the [Missouri Information for Community Assessment](https://healthapps.dhss.mo.gov/MoPhims/MICAHome). The following four categories were used, following recent [Kaiser Family Foundation](https://www.kff.org/global-health-policy/press-release/about-4-in-10-adults-in-the-u-s-are-at-greater-risk-of-developing-serious-illness-if-infected-with-coronavirus-due-to-age-or-underlying-health-conditions/) analyses of COVID-19 risk factors: Cancer, COPD, Diabetes, and Heart Disease. Mortality counts for each category were pooled for 2014-2018 and then summed to create a single measure of mortality for the entire population and for those 65 and over, expressed as a rate per 1,000.

<br>

## 65 and Over
### Rate per 1,000 Individuals
This map presents the rate of individuals 65 and over per 1,000 individuals in each ZCTA:

```{r over65, echo=FALSE, out.width = '100%'}
# create color palette
pal <- colorNumeric("Blues", over65$over65_rate)

over65 %>%
  leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    color = "#444444", 
    weight = 1, 
    opacity = 1.0, 
    smoothFactor = 0.5,
    fillOpacity = 0.5,
    fillColor = ~pal(over65_rate),
    highlightOptions = highlightOptions(color = "white", weight = 2, bringToFront = TRUE),
    popup = paste("<b>Zip Code:</b> ", over65$GEOID, "<br>",
                  "<b>Total Population:</b> ", over65$total, "<br>",
                  "<b>65 and Over Population:</b> ", over65$over65, "<br>",
                  "<b>Rate per 1,000:</b> ", round(over65$over65_rate, digits = 2))) %>%
    addLegend(pal = pal, values = ~over65_rate, opacity = .5, title = "65 and Over Rate")
```

<br>

### Mortality Rate
This map presents the mortality rate for individuals 65 and over:

```{r over65-mortality, echo=FALSE, out.width = '100%'}
# create color palette
pal <- colorNumeric("Reds", over65$mortality_rate)

over65 %>%
  leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    color = "#444444", 
    weight = 1, 
    opacity = 1.0, 
    smoothFactor = 0.5,
    fillOpacity = 0.5,
    fillColor = ~pal(mortality_rate),
    highlightOptions = highlightOptions(color = "white", weight = 2, bringToFront = TRUE),
    popup = paste("<b>Zip Code:</b> ", over65$GEOID, "<br>",
                  "<b>65 and Over Population:</b> ", over65$over65, "<br>",
                  "<b>Pooled Mortality Count:</b> ", over65$mortality, "<br>",
                  "<b>Mortality Rate per 1,000:</b> ", round(over65$mortality_rate, digits = 2))) %>%
    addLegend(pal = pal, values = ~mortality_rate, opacity = .5, title = "Mortality Rate")
```

<br>

### Bi-variate Map
This map compares mortality and population rates for those 65 and over. Areas that are dark violent have both high mortality rates and a high proportion of individuals 65 and over.

```{r bivariate-legend, echo=FALSE, out.width = '50%'}
bi_legend(pal = "DkViolet", dim = 3,
                    xlab = "65 and Over Mortality Rate",
                    ylab = "65 and Over Population Rate",
                    size = 16)
```



```{r bivariate-over65, echo=FALSE, out.width = '100%'}
# create palette
pal_df <- bi_class(over65, x = mortality_rate, y = over65_rate, style = "quantile", dim = 3)

pal <- tibble(
  bi_class = c("1-1", "1-2", "1-3",
               "2-1", "2-2", "2-3",
               "3-1", "3-2", "3-3"),
  fill = c("#CABED0", "#89A1C8", "#4885C1",
           "#BC7C8F", "#806A8A", "#435786",
           "#AE3A4E", "#77324C", "#3F2949"),
  name = c("lower over 65 mortality rate, lower over 65 population rate",
           "lower over 65 mortality  rate, moderate over 65 population rate",
           "lower over 65 mortality rate, higher over 65 population rate",
           "moderate over 65 mortality rate, lower over 65 population rate",
           "moderate over 65 mortality rate, moderate over 65 population rate",
           "moderate over 65 mortality rate, higher over 65 population rate",
           "higher over 65 mortality rate, lower over 65 population rate",
           "higher over 65 mortality rate, moderate over 65 population rate",
           "higher over 65 mortality rate, higher over 65 population rate")
  )

pal_df <- left_join(pal_df, pal, by = "bi_class")
pal_df <- mutate(pal_df, bi_class = as.factor(bi_class))

factpal <- colorFactor(pal$fill, pal_df$bi_class)

pal_df %>%
  leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    color = "#444444", 
    weight = 1, 
    opacity = 1.0, 
    smoothFactor = 0.5,
    fillOpacity = 0.7,
    fillColor = ~factpal(bi_class),
    popup = paste("<b>Zip Code:</b> ", pal_df$GEOID, "<br>",
                  "<b>65 and Over Rate per 1,000:</b> ", round(over65$over65_rate, digits = 2), "<br>",
                  "<b>Mortality Rate per 1,000:</b> ", round(over65$mortality_rate, digits = 2), "<br>",
                  "<b>Category:</b> ", pal_df$name))
```

<br>

## All Individuals
### Poverty Rate per 1,000 Individuals
This map presents the poverty rate for all individuals in each ZCTA:

```{r poverty, echo=FALSE, out.width = '100%'}
# create color palette
pal <- colorNumeric("Greens", all$poverty_rate)

all %>%
  leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    color = "#444444", 
    weight = 1, 
    opacity = 1.0, 
    smoothFactor = 0.5,
    fillOpacity = 0.5,
    fillColor = ~pal(poverty_rate),
    highlightOptions = highlightOptions(color = "white", weight = 2, bringToFront = TRUE),
    popup = paste("<b>Zip Code:</b> ", all$GEOID, "<br>",
                  "<b>Total Population:</b> ", all$total, "<br>",
                  "<b>Population Below Poverty Line:</b> ", all$poverty, "<br>",
                  "<b>Rate per 1,000:</b> ", round(all$poverty_rate, digits = 2))) %>%
    addLegend(pal = pal, values = ~poverty_rate, opacity = .5, title = "Poverty Rate")
```

<br>

### Mortality Rate
This map presents the mortality rate for all individuals:

```{r all-mortality, echo=FALSE, out.width = '100%'}
# create color palette
pal <- colorNumeric("Oranges", all$mortality_rate)

all %>%
  leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    color = "#444444", 
    weight = 1, 
    opacity = 1.0, 
    smoothFactor = 0.5,
    fillOpacity = 0.5,
    fillColor = ~pal(mortality_rate),
    highlightOptions = highlightOptions(color = "white", weight = 2, bringToFront = TRUE),
    popup = paste("<b>Zip Code:</b> ", all$GEOID, "<br>",
                  "<b>Total Population:</b> ", all$total, "<br>",
                  "<b>Pooled Mortality Count:</b> ", all$mortality, "<br>",
                  "<b>Mortality Rate per 1,000:</b> ", round(all$mortality_rate, digits = 2))) %>%
    addLegend(pal = pal, values = ~mortality_rate, opacity = .5, title = "Mortality Rate")
```

<br>

### Bi-variate Map
This map compares mortality and poverty rates for the entire population of each ZCTA Areas that are dark violent have both high mortality rates and a high poverty rate.

```{r bivariate-legend2, echo=FALSE, out.width = '50%'}
bi_legend(pal = "DkViolet", dim = 3,
                    xlab = "Mortality Rate",
                    ylab = "Poverty Rate",
                    size = 16)
```

```{r bivariate-all, echo=FALSE, out.width = '100%'}
# create palette
pal_df <- bi_class(all, x = mortality_rate, y = poverty_rate, style = "quantile", dim = 3)

pal <- tibble(
  bi_class = c("1-1", "1-2", "1-3",
               "2-1", "2-2", "2-3",
               "3-1", "3-2", "3-3"),
  fill = c("#CABED0", "#89A1C8", "#4885C1",
           "#BC7C8F", "#806A8A", "#435786",
           "#AE3A4E", "#77324C", "#3F2949"),
  name = c("lower mortality rate, lower poverty rate",
           "lower mortality  rate, moderate poverty rate",
           "lower mortality rate, higher poverty rate",
           "moderate mortality rate, lower poverty rate",
           "moderate mortality rate, moderate poverty rate",
           "moderate mortality rate, higher poverty rate",
           "higher mortality rate, lower poverty rate",
           "higher mortality rate, moderate poverty rate",
           "higher mortality rate, higher poverty rate")
  )

pal_df <- left_join(pal_df, pal, by = "bi_class")
pal_df <- mutate(pal_df, bi_class = as.factor(bi_class))

factpal <- colorFactor(pal$fill, pal_df$bi_class)

pal_df %>%
  leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    color = "#444444", 
    weight = 1, 
    opacity = 1.0, 
    smoothFactor = 0.5,
    fillOpacity = 0.7,
    fillColor = ~factpal(bi_class),
    popup = paste("<b>Zip Code:</b> ", pal_df$GEOID, "<br>",
                  "<b>Poverty Rate per 1,000:</b> ", round(pal_df$poverty_rate, digits = 2), "<br>",
                  "<b>Mortality Rate per 1,000:</b> ", round(pal_df$mortality_rate, digits = 2), "<br>",
                  "<b>Category:</b> ", pal_df$name))
```